### 2015-10-01 pset1/paper.tex
asdf

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fullpage}
\title{6.867 Problem Set 1}
\date{October 1, 2015}

\begin{document}
\maketitle
\section{Gradient Descent}
Gradient descent is a very popular numerical algorithm for finding the minimum value of a function. The algorithm works by starting witha guessed input value, and iterates towards the minimum. At each step, the algorithm computes the gradient at its input value, and shifts its guessed input along the direction opposite the gradient. When the difference between evaluations of the function at two consecutive input values falls below some convergence criterion, the algorithm stops, and the last input value is returned as the location of the minimum.

\subsection{Implementation}
We implemented gradient descent in 25 lines of Python, using the numpy library. 
\end{document}

### 2015-10-01 pset1/lad.py

@@ -1,7 +1,6 @@
 import numpy as np
 import gd
 import lbfr
-import parker_rr as rr
 from hw1_res import homework1
 
 
@@ -29,7 +28,13 @@ X_test = data_test[0].reshape(data_test[0].size)
 Y_test = data_test[1].reshape(data_test[1].size)
 
 mins = {}
+asdf = 0
 for M in range(10):
   for lam in [2 ** x for x in np.arange(-3, 6, 0.5)]:
     f = obj_func(lbfr.compute_Phi_poly(X_train, M), Y_train, lam)
-    
+    res = gd.gd(f, np.zeros(M+1), 0.001, 0.001, max_iters = 10000)
+    Theta = res[0]
+    mins[(M,lam)] = f(Theta)
+    asdf += 1
+    print asdf
+print mins

### 2015-10-01 pset1/rr.py
RR

@@ -7,20 +7,21 @@ def ridge_regression(x, y, l, M):
 	matrix_x = np.zeros(shape = (len(x), M + 1))
 	for i in xrange(len(x)):
 		matrix_x[i] = [x[i] ** j for j in xrange(M + 1)]
-	print matrix_x
 	identity = l * np.identity(M + 1)
 	matrix_y = y[:, np.newaxis]
 	return np.dot(np.dot(inv(np.dot(matrix_x.T, matrix_x) + identity), matrix_x.T), matrix_y)
 
 xdata = np.array([0.000000, 0.111111, 0.222222, 0.333333, 0.444444, 0.555556, 0.666667, 0.777778, 0.888889, 1.00000])
 ydata = np.array([0.349486, 0.830839, 1.007332, 0.971507, 0.133066, 0.166823, -0.848307, -0.445686, -0.563567, 0.261502])
-l = 1
+l = 0.00000001
 M = 10
 W = [i[0] for i in  ridge_regression(xdata, ydata, l, M)]
 
-print W
 rx = linspace(0, 1, 200)
-ry = 0.44445594686119838 -0.32704035116221264 * rx -0.30870137298037709 * (rx ** 2) -0.16092522705038409 * (rx ** 3) -0.020195360919547531 * (rx ** 4) + 0.092642816811854004 * (rx ** 5)
+
+ry = 0
+for i in xrange(len(W)):
+	ry += W[i] * (rx ** i)
 plot(rx, ry)
 
 x = [0.000000, 0.111111, 0.222222, 0.333333, 0.444444, 0.555556, 0.666667, 0.777778, 0.888889, 1.00000]

### 2015-10-01 pset1/paper.tex
most of writeup for part 1

@@
 The algorithm works by starting witha guessed input value, and iterates towards the minimum.
-At each step, the algorithm computes the gradient at its input value, and shifts its guessed input along the direction opposite the gradient.
+At each step, the algorithm computes the gradient at its input value, and shifts its guessed input along the direction opposite the gradient (the amount shifted is proportional of a parameter to the algorithm called the step size).
 When the difference between evaluations of the function at two consecutive input values falls below some convergence criterion, the algorithm stops, and the last input value is returned as the location of the minimum.
 
-\subsection{Implementation}
+\subsection{Implementation and Testing}
 We implemented gradient descent in 25 lines of Python, using the numpy library.
+13 of these lines were an implementation of a function to numerically estimate the gradient of a function.
+After verifying the validity of the gradient function on a few analytically simple functions, we tested gradient descent on the following two functions:
+\begin{align*}
+f_1(x_1, x_2) &= (x_1-2)^2 + (x_2-2)^2\\
+f_2(x_1, x_2) &= x_1^2 + x_2^2 + 4\sin (x_1 + x_2)
+\end{align*}
+$f_1$ has a unique minimum at $(x_1, x_2) = (2,2)$, whereas $f_2$ has multiple local minima: they are at $(-0.626, -0.626)$ and $(1.798, 1.798)$, the former of which is the global minimum.
+(There is a third point at which the gradient of $f_2$ is zero, but that point is a saddle point.)
+
+When testing with $f_1$, we found that for values of the step size between 0.1 and 0.7 and starting guesses $-10^{9} \le x_1, x_2 \le 10^{9}$, the algorithm converged to the minimum of $(2,2)$ fairly quickly (under 200 iterations), even with convergence criteria as small as $10^{-8}$.
+For significantly larger starting guesses, our gradient algorithm failed due to rounding errors, and for much smaller step sizes, our algorithm needed more steps to converge.
+For large step sizes (greater than $0.8$), the algorithm would actually take longer, or if the step size is very large, fail to converge altogether.
+This is due to the descent ``overshooting the minimum" on each iteration of the algorithm (i.e.
+the iteration moves the input point in the direction of the minimum, but moves much too far).
+
+Testing $f_2$ led to some similar results.
+We found that step sizes between 0.1 and 0.3 and starting values $-10^{9} \le x_1, x_2 \le 10^9$ allowed the algorithm to converge quickly for a convergence criterion of $10^{-8}$.
+Adjusting the step size outside of this range led to similar behavior in that the algorithm would not converge, but interestingly, for this function, the algorithm would often get stuck on the $x_1 = x_2$ line.
+This is perhaps due to the fact that $f_2$ is a symmetric function.
+
+However, when testing $f_2$, we noticed that sometimes the algorithm would terminate in one minimum, and sometimes it would terminate in the other.
+(We were unable to produce a nontrivial example of the algorithm terminating at the saddle point.) The general pattern for starting guesses near the origin ($-10 \le x_1, x_2 \le 10$) is that starting guesses will converge to the minimum nearest them.
+This pattern was not followed exactly when the starting guess was significantly farther away from the origin, but it was still roughly true.
+This behavior is unsurprising; one would expect most functions to be contoured such that the gradient at most locations ``points most closely" to the closest local minimum.
 \end{document}

### 2015-10-01 pset1/paper.tex
problem 1 writeup

@@
 
-When testing with $f_1$, we found that for values of the step size between 0.1 and 0.7 and starting guesses $-10^{9} \le x_1, x_2 \le 10^{9}$, the algorithm converged to the minimum of $(2,2)$ fairly quickly (under 200 iterations), even with convergence criteria as small as $10^{-8}$.
+When testing with $f_1$, we found that for values of the step size between 0.1 and 0.7 and starting guesses $-10^{9} \le x_1, x_2 \le 10^{9}$, the algorithm converged to the minimum of $(2,2)$ fairly quickly (under 200 iterations), even with convergence criteria as small as $10^{-16}$.
 For significantly larger starting guesses, our gradient algorithm failed due to rounding errors, and for much smaller step sizes, our algorithm needed more steps to converge.
@@
 Testing $f_2$ led to some similar results.
-We found that step sizes between 0.1 and 0.3 and starting values $-10^{9} \le x_1, x_2 \le 10^9$ allowed the algorithm to converge quickly for a convergence criterion of $10^{-8}$.
+We found that step sizes between 0.1 and 0.3 and starting values $-10^{9} \le x_1, x_2 \le 10^9$ allowed the algorithm to converge quickly for a convergence criterion of $10^{-16}$.
 Adjusting the step size outside of this range led to similar behavior in that the algorithm would not converge, but interestingly, for this function, the algorithm would often get stuck on the $x_1 = x_2$ line.
@@
 However, when testing $f_2$, we noticed that sometimes the algorithm would terminate in one minimum, and sometimes it would terminate in the other.
-(We were unable to produce a nontrivial example of the algorithm terminating at the saddle point.) The general pattern for starting guesses near the origin ($-10 \le x_1, x_2 \le 10$) is that starting guesses will converge to the minimum nearest them.
-This pattern was not followed exactly when the starting guess was significantly farther away from the origin, but it was still roughly true.
-This behavior is unsurprising; one would expect most functions to be contoured such that the gradient at most locations ``points most closely" to the closest local minimum.
+(We were unable to produce a nontrivial example of the algorithm terminating at the saddle point.) The general pattern was that starting guesses would converge to the minimum nearest them.
+This behavior is unsurprising; one would expect most functions to be contoured such that the gradient at most locations is roughly in the direction of the closest local minimum.
+
+\subsection{Comparison with more sophisticated algorithms}
+We compared our implementation of gradient descent with a more sophisticated algorithm, the \begin{tt}fmin\_bfgs\end{tt} function in the \begin{tt}scipy.optimize\end{tt} package.
+We used both \begin{tt}fmin\_bfgs\end{tt} and our own implementation of gradient descent to minimize $f_1$ and $f_2$ from a variety of starting points.
+With our own gradient descent algorithm, we used $10^{-16}$ as the convergence criterion for both minimizations, and step sizes of 0.4 and 0.2 for $f_1$ and $f_2$, respectively.
+With \begin{tt}fmin\_bfgs\end{tt}, we simply used default values.
+
+We found that \begin{tt}fmin\_bfgs\end{tt} usually required equal or more function evaluations, but fewer gradient evaluations, than unmodified gradient descent when minimizing $f_1$, while achieving similar accuracy.
+For example, with starting guess $(10^5, -10^5)$, gradient descent performed 20 function calls and 20 gradient calls, whereas \begin{tt}fmin\_bfgs\end{tt} performed 28 function calls and 7 gradient calls.
+The same held true for $f_2$ when the starting guess roughly satisfied $x_1, x_2 \ge 3$ and $\max (x_1, x_2) \le 1.1\min (x_1, x_2)$.
+However, when this was not true, unmodified gradient descent required many more steps than \begin{tt}fmin\_bfgs\end{tt}: unmodified gradient descent would usually terminate after performing 150-200 function and gradient calls, whereas \begin{tt}fmin\_bfgs\end{tt} would typically perform 30-50 function calls and 10-15 gradient calls (we were unable to determine the cause of this behavior).
+The two algorithms also sometimes ended up at different minima; for example, when starting from $(x_1, x_2) = (-10, -10)$, gradient descent terminated at the global minimum, while \begin{tt}fmin\_bfgs\end{tt} terminated at the other local minimum.
+
 \end{document}

### 2015-10-02 pset1/rr.py
Plotting

@@ -1,34 +1,118 @@
 import numpy as np
 from numpy.linalg import inv
 from pylab import *
+import shelve
 
 # x is numpy row array, y is numpy row array, l is lambda, M is degree of the basis
-def ridge_regression(x, y, l, M):
+def ridge_regression(x, y, L, M):
 	matrix_x = np.zeros(shape = (len(x), M + 1))
 	for i in xrange(len(x)):
 		matrix_x[i] = [x[i] ** j for j in xrange(M + 1)]
-	identity = l * np.identity(M + 1)
+	identity = np.identity(M + 1)
 	matrix_y = y[:, np.newaxis]
-	return np.dot(np.dot(inv(np.dot(matrix_x.T, matrix_x) + identity), matrix_x.T), matrix_y)
+	return np.dot(np.dot(inv(np.dot(matrix_x.T, matrix_x) + L * identity), matrix_x.T), matrix_y)
+
+def produce_plot(xdata, ydata, L, M, name):
+	W = [i[0] for i in ridge_regression(xdata, ydata, L, M)]
+
+	rx = linspace(min(xdata), max(xdata), 200)
+	ry = 0
+	for i in xrange(len(W)):
+		ry += W[i] * (rx ** i)
+
+	plot(rx, ry)
+	plot(xdata, ydata, 'bo', label = 'sampled')
+	grid(True)
+	savefig(name)
+	clf()
+
+def error(x, y, L, theta, M):
+	matrix_x = np.zeros(shape = (len(x), M + 1))
+	for i in xrange(len(x)):
+		matrix_x[i] = [x[i] ** j for j in xrange(M + 1)]
+	matrix_y = y[:, np.newaxis]
+	yhat = np.dot(matrix_x, theta)
+	return np.linalg.norm(yhat - matrix_y)**2 + L * (np.dot(theta.T, theta))
+
+def getData(name):
+    data = loadtxt(name)
+    X = data[0:1]
+    Y = data[1:2]
+    return X, Y
 
 xdata = np.array([0.000000, 0.111111, 0.222222, 0.333333, 0.444444, 0.555556, 0.666667, 0.777778, 0.888889, 1.00000])
 ydata = np.array([0.349486, 0.830839, 1.007332, 0.971507, 0.133066, 0.166823, -0.848307, -0.445686, -0.563567, 0.261502])
-l = 0.00000001
-M = 10
-W = [i[0] for i in  ridge_regression(xdata, ydata, l, M)]
-
[…]

### 2015-10-02 pset1/paper.tex
wrote 1, part of 2

@@
 
+\section{Linear Basis Function Regression}
+Given a feature matrix $\Phi$, a weight vector $\theta$, and an output vector $Y$, we define the sum-of-squares error (SSE) of $\theta$ as
+$$(\Phi\theta - Y)^T(\Phi\theta - Y).$$
+The particular vector $\theta$ that minimizes this value is
+$$\theta = (\Phi^T\Phi)^{-1}\Phi^TY.$$
+The gradient of the sum-of-squares error with respect to $\theta$ is
+$$2\Phi^T(\Phi\theta - Y).$$
+We implemented functions to calculate, using the closed forms given above, both the SSE-minimizing value (i.e.
+the value with the maximum likelihood) of $\theta$ as well as the gradient of the SSE with respect to $\theta$.
+We verified the correctness of the maximum-likelihood function by checking it against data sets for which the maximum likelihood vectors were known.
+We also verified the correctness of the gradient by checking it against a numerical implementation.
+
 \end{document}

### 2015-10-02 pset1/gd.py

@@ -1,18 +1,25 @@
 import numpy as np
 import copy
-ep = 0.1
+import scipy.optimize as opt
+import matplotlib.pyplot as plt
+ep = 0.000001
 
 
-def gd(f, start, step, cvg):
+def gd(f, start, step, cvg, max_iters=100000, print_vals=False):
   start = start.astype(float)
   iters = 1
-  while iters < 100000:
+  vals = [start]
+  while iters < max_iters:
     grad = gradient(f, start)
     nextv = start - step * grad
+    vals.append(nextv)
+    if print_vals:
+      print nextv
     if abs(f(start) - f(nextv)) < cvg:
-      return nextv, iters
+      return nextv, iters, vals
     start = nextv
     iters += 1
+  return nextv, iters, vals
 
 def gradient(f, start):
   start = start.astype(float)
@@ -27,8 +34,22 @@ def gradient(f, start):
   plusv = np.array([f(startd1[i,:]) for i in range(n)])
   minusv = np.array([f(startd2[i,:]) for i in range(n)])
   return 1/(2*ep) * (plusv - minusv)
+"""
+def f1(x):
+  return (x[0] - 2) ** 2 + (x[1] - 2) ** 2
 
-def square(x):
-  return (x - 3) ** 2
+def f2(x):
+  return x[0]**2 + x[1]**2 + 4 * np.sin(x[0] + x[1])
 
-print gd(square, np.array([0]), 0.2, 0.00000001)
+guess = np.array([-10, -10])
+cvg = 0.0000000000000001
+
+print gd(f1, guess, 0.4, cvg)
+a = gd(f2, guess, 0.2, cvg)
+print a
+print opt.fmin_bfgs(f1, guess, full_output=True)
+print guess
+print opt.fmin_bfgs(f2, guess, full_output=True)
+plt.plot([x[0] for x in a[2]], [x[1] for x in a[2]])
+plt.show()
+"""

### 2015-10-02 pset1/lbfr.py

@@ -1,5 +1,6 @@
 import numpy as np
 from hw1_res import homework1
+import pylab
 import gd
 
 
@@ -42,7 +43,9 @@ for M in [0,1,3,9]:
   Theta = max_likely(X,Y,M)
   Theta2 = np.zeros(Theta.size)
   Phi = compute_Phi_poly(X,M)
+
   print "-----------------"
+  print Theta
   print sse(Theta, Phi, Y)
   print sse_der(Theta2, Phi, Y)
   print sse_der_2(Theta2, Phi, Y)

### 2015-10-02 pset1/paper.tex
done with 2

@@
 
+\subsection{Minimizing SSE with gradient descent}
+We attempted to find maximum-likelihood values of $\theta$ using unmodified gradient descent and \begin{tt}fmin\_bfgs\end{tt}.
+Unmodified gradient descent ended up performing very poorly on this problem; with $M=3$ and a convergence criteria of $10^{-8}$, 100,000 iterations (each iteration including one function call and one gradient call) were not enough for the algorithm to converge with a step size of 0.01.
+The best-performing step size we found was 0.06, which converged in 49,003 iterations.
+Larger step sizes (as small as 0.065) ended up diverging and giving us no useful results.
+Reducing the convergence criteria to $10^{-5}$ (at which point there was a visible discrepancy between the maximum-likelihood polynomials produced by the closed-form formula and that produced by gradient descent) did not impact performance significantly; with a step size of 0.06, the algorithm still required approximately 30,000 iterations to converge.
+Adjusting the starting guess did not significantly affect the performance of the algorithm.
+
+\begin{tt}fmin\_bfgs\end{tt} performed much better on this problem.
+The algorithm required onnly 162 function calls and 27 gradient calls (again, with the $M=3$ case) to converge to a solution (which was visually indistinguishable from the optimal solution found by the closed-form formula).
+Plots depicting our results can be found in Figure 1.
+
+\begin{figure}
+\includegraphics[scale=0.4]{figure2_1.png}
+\includegraphics[scale=0.4]{figure2_2.png}
+\caption{The maximum-likelihood polynomials ($M=3$) produced by gradient descent.
+The sin function used to generate the data is in green, the polynomial produced by the closed-form formula in red, and the polynomial produced by gradient descent in blue.
+Note that in the left figure, the red curve is almost invisible, as it is mostly covered by the blue curve.
+The figure on the left corresponds to a convergence criterion of $10^{-8}$, while the one on the right corresponds to a convergence criterion of $10^{-5}$.
+The \texttt{fmin\_bfgs} function produced a figure similar to the one on the left.}
+\end{figure}
+
+Because the data set that we used was generated using an underlying sin function, we investigated the behavior of the algorithm when trigonometric functions were used to generate $\Phi$.
+We initially expected that the algorithm would generate a fit that was closer to the original sin function, but it actually did not.
+The result of this can be seen in Figure 2.
+Attempting to fit data in this way without knowing how it was generated also carries some disadvantages.
+Any function that we generate with trigonometric basis functions must necessarily be periodic with period 1, so the fit may not generalize well to values of $x$ outside the interval supplied.
+Furthermore, trigonometric functions will have a difficult time characterizing simple relationships, such as linear correlations.
+
+\begin{figure}
+\includegraphics[scale=0.4]{figure2_3.png}
+\caption{The maximum-likelihood function produced using trigonometric basis functions ($M=3$).
+This function (blue) does not actually appear to be more faithful to the original sin function (green) than the maximum-likelihood degree-3 polynomial (red).}
+\end{figure}
 \end{document}

### 2015-10-02 pset1/lbfr.py

@@ -2,6 +2,7 @@ import numpy as np
 from hw1_res import homework1
 import pylab
 import gd
+import scipy.optimize as opt
 
 
 def compute_Phi_poly(X, M):
@@ -15,6 +16,7 @@ def compute_Phi_trig(X, M):
   Phi = np.zeros((X.size, M+1))
   for i in range(M+1):
     Phi[:,i] = np.sin(2 * i * np.pi * X)
+  Phi[:,0] = np.ones(Phi[:,0].size)
   return Phi
 
 
@@ -36,19 +38,29 @@ def sse_der_2(Theta, Phi, Y):
   f = lambda x: sse(x, Phi, Y)
   return gd.gradient(f, Theta)
 
+
 data = homework1.getData('curvefitting.txt')
 X = data[0].reshape(data[0].size)
 Y = data[1].reshape(data[1].size)
-for M in [0,1,3,9]:
+pylab.plot(X, Y, 'bo')
+for M in [3]:
   Theta = max_likely(X,Y,M)
-  Theta2 = np.zeros(Theta.size)
-  Phi = compute_Phi_poly(X,M)
+  Theta2 = np.ones(Theta.size) * (10 ** 5)
+  Phi = compute_Phi_trig(X,M)
+  f = lambda theta: sse(theta, Phi, Y)
+  fprime = lambda theta: sse_der(theta, Phi, Y)
+  asdf = opt.fmin_bfgs(f, Theta2, full_output=True)
+  Theta2 = asdf[0]
+  print Theta2
+  Xlin = np.linspace(0,1,200)
+  Philin = compute_Phi_poly(Xlin, M)
+  Phitrig = compute_Phi_trig(Xlin, M)
+  Ylin_poly = np.dot(Philin, Theta)
+  Ylin_sin = np.array([np.sin(2 * x * np.pi) for x in Xlin])
+  Ylin_poly_2 = np.dot(Phitrig, Theta2)
+  if M == 3:
+    pylab.plot(Xlin, Ylin_poly, 'r')
+    pylab.plot(Xlin, Ylin_sin, 'g')
+    pylab.plot(Xlin, Ylin_poly_2, 'b')
+pylab.show()
 
-  print "-----------------"
-  print Theta
-  print sse(Theta, Phi, Y)
-  print sse_der(Theta2, Phi, Y)
-  print sse_der_2(Theta2, Phi, Y)
-
-
-print compute_Phi_trig(X, 3)

### 2015-10-02 pset1/paper.tex
done writing 1 2 4

@@
 The algorithm required onnly 162 function calls and 27 gradient calls (again, with the $M=3$ case) to converge to a solution (which was visually indistinguishable from the optimal solution found by the closed-form formula).
-Plots depicting our results can be found in Figure 1.
+Plots depicting our results can be found in Figure \ref{p2figure1}.
 
 \begin{figure}
+\label{p2figure1}
 \includegraphics[scale=0.4]{figure2_1.png}
@@
 We initially expected that the algorithm would generate a fit that was closer to the original sin function, but it actually did not.
-The result of this can be seen in Figure 2.
+The result of this can be seen in Figure \ref{p2figure2}.
 Attempting to fit data in this way without knowing how it was generated also carries some disadvantages.
@@
 \begin{figure}
+\label{p2figure2}
 \includegraphics[scale=0.4]{figure2_3.png}
@@
 \end{figure}
+
+\section{Least Absolute Deviation}
+Alternatively to minimizing the SSE in a regression problem, we can instead try to minimize the sum of the absolute errors, which is a method called Least Absolute Deviation (i.e.
+LAD).
+In this section we investigate the behavior of this method of optimization.
+
+We repeated our experiment from section 3 in this section, fitting weight vectors for values of $M$ between 0 and 9 (inclusive), and values of $\lambda$ of the form $2^{k/2}$ for integers $-3 \le k \le 11$.
+We found the cross-validation error for each weight vector, and discovered that $(M, \lambda) = (2, 2^{-3})$ resulted in the lowest cross-validation error, while $(M, \lambda) = (6, 2^{-1})$ resulted in the highest.
+However, the best and worst CV errors differed by only one order of magnitude, which surprised us.
+We believe this is due to an affinity towards median behavior of this method, as the value with the least absolute error with respect to a set of values is the median of that set.
+The best and worst fits are depicted in Figure \ref{p4figure}.
+
+\begin{figure}
+\label{p4figure}
+\includegraphics[scale=0.4]{figure4_3.png}
+\includegraphics[scale=0.4]{figure4_4.png}
+\caption{The fits with the lowest (left) and highest (right) cross-validation errors.
+The left fit used the values $M=2, \lambda = 2^{-3}$, while the right fit used the values $M=6, \lambda = 2^{-1}$.
+The fitted polynomial is shown in red, with the training set, cross-validation set, and test set shown in blue, red, and green, respectively.}
+\end{figure}
+
+Upon looking at all of the generated data, it appeared that the value of $\lambda$ had little consequence on the cross-validation error, and that there was at best a slight positive relationship between the value of $M$ and the cross-validation error.
+It thus seems that using ridge regression has little consequence when using LAD.
+Ridge regression seems more suitable in situations where overfitting may be a problem, and because LAD demonstrates affinity for median-like behavior, it is unlikely to overfit.
+We believe that LAD is a suitable method to use when other methods, such as ridge regression, are insufficient for solving the overfitting problem, or when finding optimal values of $M$ and $\lambda$ is impractical (many values to try, large datasets that take a long time to fit, etc.).
 \end{document}

### 2015-10-02 pset1/lad.py

@@ -1,7 +1,8 @@
 import numpy as np
-import gd
+import scipy.optimize as opt
 import lbfr
 from hw1_res import homework1
+import pylab
 
 
 def compute_lad(Phi, Theta, Y, lam=0):
@@ -32,9 +33,29 @@ asdf = 0
 for M in range(10):
   for lam in [2 ** x for x in np.arange(-3, 6, 0.5)]:
     f = obj_func(lbfr.compute_Phi_poly(X_train, M), Y_train, lam)
-    res = gd.gd(f, np.zeros(M+1), 0.001, 0.001, max_iters = 10000)
+    res = opt.fmin_bfgs(f, np.zeros(M+1), full_output=True)
     Theta = res[0]
-    mins[(M,lam)] = f(Theta)
-    asdf += 1
-    print asdf
-print mins
+    mins[(M,lam)] = compute_lad(lbfr.compute_Phi_poly(X_cv, M), Theta, Y_cv, lam)
+
+    if M == 6 and lam > 0.4 and lam < 0.55:
+      Xlin = np.linspace(-3, 2, 500)
+      Philin = lbfr.compute_Phi_poly(Xlin, M)
+      Ylin_poly = np.dot(Philin, Theta)
+      pylab.plot(Xlin, Ylin_poly, 'r')
+      pylab.plot(X_train, Y_train, 'bo')
+      pylab.plot(X_cv, Y_cv, 'ro')
+      pylab.plot(X_test, Y_test, 'go')
+min_k = None
+for k in mins:
+  if min_k == None or mins[k] < mins[min_k]:
+    min_k = k
+plist = []
+for k in mins:
+  plist.append((mins[k], k))
+plist.sort()
+for i in plist:
+  print i
+print min_k
+print mins[min_k]
+print X_train
+pylab.show()

### 2015-10-02 pset1/lbfr.py

@@ -38,7 +38,7 @@ def sse_der_2(Theta, Phi, Y):
   f = lambda x: sse(x, Phi, Y)
   return gd.gradient(f, Theta)
 
-
+"""
 data = homework1.getData('curvefitting.txt')
 X = data[0].reshape(data[0].size)
 Y = data[1].reshape(data[1].size)
@@ -63,4 +63,4 @@ for M in [3]:
     pylab.plot(Xlin, Ylin_sin, 'g')
     pylab.plot(Xlin, Ylin_poly_2, 'b')
 pylab.show()
-
+"""

### 2015-10-02 pset1/paper.tex
asdf

@@
 
-We repeated our experiment from section 3 in this section, fitting weight vectors for values of $M$ between 0 and 9 (inclusive), and values of $\lambda$ of the form $2^{k/2}$ for integers $-3 \le k \le 11$.
+We repeated our experiment from section 3 in this section, fitting weight vectors for values of $M$ between 0 and 9 (inclusive), and values of $\lambda$ of the form $2^{k/2}$ for integers $-6 \le k \le 11$.
 We found the cross-validation error for each weight vector, and discovered that $(M, \lambda) = (2, 2^{-3})$ resulted in the lowest cross-validation error, while $(M, \lambda) = (6, 2^{-1})$ resulted in the highest.

### 2015-10-02 pset1/paper.tex
done

--- 2015-10-02 pset1/paper.tex
@@
 
+\section{Ridge Regression}
+With regression on certain datasets (for example, one with many features but not enough data), we often run into the issue of overfitting.
+In other words, our model may fit the training set very well but fails to generalize.
+We can address this using regularization, which includes the squared norm of the weight vector as a penalty, scaled by a factor $\lambda$ that induces bias to counter variance.
+
+\subsection{Implementation and Testing}
+Using the closed form $\theta = (\phi ^ {T} \phi + \lambda I)^{-1} \phi ^ T Y$, we implemented ridge regression in Python, using the numpy library.
+Then, using the simple data from Bishop's Figure 1.4, we ran ridge regression using $M$ ranging from $0$ to $10$ and $\lambda$ from $0, 0.0001, 0.0002 \dots, 5.24288$.
+We noted that for larger values of $M$, while the model fit the data well, it did not generalize well past the given range of values, suggesting that small to intermediate $M$ values may fit the data better.
+Furthermore, for larger values of $M$, changing $\lambda$ slightly had a greater effect as the elements of $\theta$ were fairly large, so penalizing the weight vector incurred a large cost.
+Using $\lambda = 0$ as a baseline, we were able to witness the effects of regularization in smoothing out curves.
+Considering $\lambda$ in particular, we note that large lambda values result in high bias and thus underfitting, while low lambda values (approximately 0) result in high variance and thus overfitting, in line with our previous assumptions.
+
+\begin{figure}
+\label{p3figure1}
+\includegraphics[scale=0.4]{bishop7.png}
+\caption{Ridge regression with $M = 7, \lambda = 0.00128$}
+\end{figure}
+
+\subsection{Model Selection}
+Regarding the process of model selection, we initially began with $M$ again ranging from $0$ to $10$ and 200 values for $\lambda$ ranging from $10 ^ (-8)$ to $1.726$, generated by continually multiplying $1.1$ to an initial $\lambda$.
+Then, using the training data set, we were able to generate $\theta$ values for all combinations of $M$ and $\lambda$.
+We then tested each of these $\theta$ values on the validation set, and computed the error generated.
+Then, using the models with the smallest error produced, we applied them on the testing set to get our final results.
+\\ \\
+From testing our $\theta$'s on the validation set, we were able to secure $M = 4, \lambda = 0.65536; M = 3, \lambda = 0;$ and $M = 3, \lambda = 10 ^ {-5}$ with respective errors $3.208586, 3.6360459, 3.636073$.
+Considering the testing set, we arrived at $M = 3, \lambda = 10 ^ {-5}$ as our model, with the lowest error of $0.2496965$.
+Graphs have been produced below.
+
+\subsection{BlogFeedback}
+Ridge regression is especially suitable here as the blog dataset is one with many features (280) in fact, but not too much data (most of the identified values being $0$).This was slightly different from our previous model selection process as we are using a simple linear model here.
+Thus, as the focus was on finding the appropriate value for $\lambda$, we used an exhaustive approach and considered values in range $10 ^ {-5}$ to $1$ in increments of $10 ^ (-5)$.
+Proceeding through identifying $\theta$ based on the training dat set, validating the model with the validation dataset, and finally testing with the testing dataset, we arrived at $\lambda = 0.0091315$, with a surprisingly high error of $8051727.304087$.
+
 \section{Least Absolute Deviation}
